---
title: "Workshop 8 TEMPLATE"
subtitle: "9am"
author: "Jocelyn Magner"
date: "5/22/26"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
       sheet = "First Harvest")
  )

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col,
  "Stipa pulchra" = stipul_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col)


# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating new objecct 
drought_exp_clean <- drought_exp |>
  # cleaning names to all lowercase and underscores
  clean_names() |>
  # renaming species columns
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |>
  # renaming treatment columns... why is this case match fine??
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |>
  # reordering species/set levels and choose reference and water treatment
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |>
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |>
  # keeping relevant columns
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

looking at data structure using str()
```{r}
#| label: Checking structure

str(drought_exp_clean)
```


Insert code to display 10 random rows from the data frame here.  

displaying data using slice_sample()

```{r}
#| label: display 10 rows
slice_sample(
  drought_exp_clean, #data frame to display
  n=10) # set 10 rows
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
ggplot(data= drought_exp_clean,
       mapping = aes(x= sla,
                     y= total_g))+
  geom_point()+ # add scatter
labs(x = "Specific Leaf Area (cm\U00b2 /g)", #unicode for squared
     y = "Total plant mass (g)",
     title = "As SLA increases total mass increases")

```


## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.


```{r}
ggplot(data= drought_exp_clean,
       mapping = aes(x= water_treatment,
                     y= total_g,
                     color = water_treatment))+
  geom_jitter(width = 0.1,
              height = 0)+
  
scale_color_manual(values = c("Drought stressed" = ds_col,
                   "Well watered" = ww_col))+
  
  labs(x = "Water Treatment",
       y = "Total mass (g)",
       title = "The median value for well watered plants is greater\nthan drought stressed")+
  theme(legend.position = "none")

```





## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.


```{r}
ggplot(data= drought_exp_clean,
       mapping = aes(x= species_name,
                     y= total_g,
                     color = species_name))+
  geom_jitter(width = 0.1,
              height = 0)+


  scale_color_manual(values = plant_colors)+
  
  # wrapping x axis text
  scale_x_discrete(labels = label_wrap(10))+
  labs(x = "Species",
       y = "Total mass (g)",
       title = "Mean plant mass varies per species")+
  theme(legend.position = "none")

```


# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <-lm(total_g~1,
            data = drought_exp_clean)

# model 1: all predictors
model1 <-lm(total_g ~ sla + water_treatment +species_name,
            data = drought_exp_clean)

# model 2: SLA and water treatment
model2 <- lm(total_g ~ sla + water_treatment,
           data = drought_exp_clean)

# model 3: SLA and species
model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)

# model 4: water treatment and species
model4 <- lm(total_g~ water_treatment + species_name,
             data = drought_exp_clean)

# model 5: SLA
model5 <- lm(total_g ~ sla,
             data=drought_exp_clean)

# model 6: water treatment
model6 <- lm(total_g ~ water_treatment,
             data = drought_exp_clean)

# model 7: species
model7 <- lm(total_g ~ species_name,
             data=drought_exp_clean)
```

## Model diagnostics

```{r}
#| label: check diagnostics

par(mfrow = c(2,2))
    plot(model4)

```


redisuals appear normally distributed, homoscedastic (constant variance), & no clear outliers influencing mode coefficient est


## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: calculating AIC
AICc(
  model0,
  model1,
  model2,
  model3,
  model5,
  model6,
  model7)|>
arrange(AICc)
```



Insert code to display the summary for the best model. 

model4 water treatment and species
```{r}
#| label: model summary
summary(model4)
```


- `intercept` : model prediciton for plant mass for lotus scoparius in the drought stresseed treatment.
- `water treatment`: well watered: model prediction for the dfference in lotus scop. mass between the well watered and drought stressed treatment.
- `species name penstemoncent.`: model pred for the diff btwn penstemon mass in the well watered treatment an d lotus mass in the wall watered treatment 
-`species_nameEschscholzia californica ` - model pred for the diff btwn Esc. mass in WW treat and lotus mass in WW treatment

each est diff btwen model pred btwn given species and reference treatment (Drought stressed).


## Stop and think: what does this model mean?  

**What is the best model?**

Model 4: water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

adj-r^2 72.6% of variation in plant mass is explained by a model with both of these predictors 

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Treatment influences plant mass 
Species influences plant mass. 

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.
```{r}
#| label: model predictoins

preds <- ggpredict(
  
  model4,
  terms = c("water_treatment", "species_name"))|>
  rename(water_treatment = x,
         species_name = group)
)

```



# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.
```{r}
ggplot(data = drought_exp_clean,
       mapping = aes(x = water treatment,
                     y = total_g,
                     color = water_treatment))+
  geom_jitter(height = 0,
              width = 0.1)+
  geom_pointrange(data = preds,
                  mapping = aes(x=water_treatment,
                                y=predicted,
                                ymin = conf.low,
                                ymax=conf.high)+
  facet_wrap(~species_name)+
    theme(legend.position = "none")
```


# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







